Palomar 5 Gaps

### 2024-10-21 paper/main.tex
bib is growing!

\newcommand{\paola}[1]{\textcolor{magenta}{{#1}}}
\newcommand{\salvatore}[1]{\textcolor{green}{{#1}}}

@@

  As laid out in \citet{2021PrPNP.11903865A}'s review, despite significant progress the true nature of dark matter remains elusive and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and galactic scales. A key focus today is detecting dark matter on sub-galactic scales, because at these scales different cosmological scenarios predict different lumpiness of the halos. For instance, \citet{2008ApJ...679.1260M} compared the Via Lactea simulations to some that they made themselves and showed that you can have a large variation for the number of dark matter subhalos within a galaxy. \citet{2019Galax...7...81Z} provides a review for the different proposed mechanisms of forming dark matter haloes from perturbation theory and deviations from a uniformly dense medium and also disccses techniques and limitations for modelign such physical phenomena. \citet{2019Galax...7...81Z} rightly states that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive. The possibilities for sub-galactic behaviour and structure of dark matter are vaste. However, as proposed by Carlberg in 2012, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2006ApJ...637L..29B, 2009ApJ...693.1118G}. The review of Bonaca properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{2016A&A...595A...1G}.

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{2023MNRAS.520.5225M} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose memebers have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020 reports that the Phenoix stream is dynamically consistent from a globular cluster \footnote{see bovy for a description on the difference between dwarf and cluster streams} yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times. Nicolas Martin in 2022 showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, Vitral and Bouldrini demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are over XXX cluster in the internal regional of the galaxy, a region overwhich stream finder was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, Balbinot demonstrates that internal dynamics preferentially kicks out small stars, which have lower luminosities and can make detection more difficult.

  One of the interests in studying stellar streams, is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of \paola{dark matter} haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams that yield the most information about the galactic potential. However, the use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, johnston05, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. Nonetheless, a recent goal of stellar stream astronomy was realized by \citet{2024ApJ...967...89I} constructed the most accurate model of the Milky Way to date based on a set of about 90 streams.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. Carlberg 2013 further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{2019ApJ...880...38B} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship from M et al 20XX.

  Our study focuses on the influence of globular clusters on stellar streams, specifically examining whether globular clusters could produce gaps similar to those expected from dark matter subhalos. If globular clusters do indeed create such gaps, this factor must be considered when using stellar streams to detect dark matter. An additional and unique aspect of our study is that we focus on modeling real stellar streams and not just pure simulated streams. We focus on the Palomar 5 stream, which is well-studied, has observable gaps \citep{2020ApJ...889...70B}, and is highly visible due to its position above the galactic disk. We present a result from one of our simulations in Fig.~\ref{fig:stream_on_sky}, and the following sections describe our modeling of the stream, the gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations affecting the stream.

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by Baumgardt and Vasiliev et al. We handle the uncertainties through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions and includes a covariance term between the proper motions. We repeate the whole experiment 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. However, for the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

    We convert the initial conditions from sky coordinates into a galactocentric reference frame. We use the local standard of rest and the peculiary motion of the sun as reported by Sch\"onrich 2010. The circular velocity at the solar radius was set to: $v_{\text{LSR}} = 240$~km~s$^{-1}$. The peculiar velocity of the sun was set to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$. The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from Chen et al 2001 and the galacto-radial distance was taken from Reid et al 2014. These transformations were performed using the \texttt{astropy} package. (for the next paper, I would also like to sample these uncertainties)

    For the galactic potential, we employed the second static model proposed by Pouliasis et al. (2017), which consists of a superposition of a thin disk, thick disk, and a dark matter halo. The masses and scale lengths are provided in Table 1 of Ferrone et al 2023. This model is time-independent over the course of our simulations.

    The primary methodological departure from Ferrone et al. (2023) is that now we include globular cluster interactions. Instead of treating the clusters as point masses that only experience the gravitational potential of the galaxy, we now compute N-body interactions between the clusters during the backward integration over 5 billion years. The equations of motion for the globular cluster system are thus: 
    \begin{equation}
      \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
    \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galacto-centric position, the index $i$ indicates a globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates. The index $j$ indicates the other globular cluster that are summed over where: $m_j$ is its mass, $b_j$ is its plummer scale radius, and $r_j$ is it's galacto-centric position. $\Phi$ is the axis-symmteric potential of the model used for the Milky Way.

    At the computed positions 5 Gyr ago, each globular cluster was gifted a system of 100,000 particles, sampled from a Plummer distribution to represent the stellar content of the cluster. For eaching sampling of the initial conditions, a different sampling of the positions of the 100,000 star particles was performed. This is becuase, we sampled the uncertainties on the mass in this experiment contraty to Ferrone et al 2023.

@@

    The integration was performed using a leapfrog integrator. The timestep was chosen to be sufficiently small to ensure both the conservation of energy and that the integration error over 5 Gyr was negligible compared to the scale radius of the globular clusters. Further details on the timestep selection are discussed in the appendix.

@@

    At the end of the simulations, some gaps in the stellar stream are so prominent that they can be identified visually without any sophisticated quantitative analysis. However, upon closer inspection, more subtle gaps emerge, highlighting the need for quantitative methods to systematically detect and assess these features.

    We initially sought an automated, quantitative approach from signal and image processing that could reliably detect all gaps and evaluate their significance. Unfortunately, despite extensive efforts, an ideal method remained elusive. As a result, we opted for a simpler approach: a combination of visual inspection aided by comparative analysis.

    To facilitate this process, we conducted a series of \texttt{reference} simulations. These control simulations used the same 50 sets of initial conditions but excluded interactions with globular clusters, except for the host cluster, Palomar 5. In other words, the gravitational influence of all other clusters on the star particles within the stream was omitted. By comparing the morphological differences between these control simulations and the \texttt{full} simulations (which included cluster interactions), we could more effectively identify and assess the impact of globular clusters on the stream's structure. Details on gap identification are provided in Appendix~\ref{sec:gap_detection}, and the identification of corresponding perturbing globular clusters is detailed in Appendix~\ref{sec:Perturber_Identification}.

    Comparisons between the \texttt{reference} and \texttt{full} simulations were performed in the tail coordinate system. This system allows for consistent morphological comparisons by aligning the stream with the cluster's orbit. The coordinate system is based on the work of Dehnen et al. (2004) and is detailed in Appendix~\ref{appendix:TailCoordinates}. Briefly, the x' coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead or behind the cluster. The y' coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer.

    Once the gaps and perturbers were identified, we sought to characterize the properties of the globular clusters that caused gaps and differentiate them from those that did not. Carlberg (2012) initiated the theory of gap creation in stellar streams due to gravitational flybys. This analysis was extended both analytically and numerically by Erkal et al. (2015), who studied uniform-density streams on circular orbits. Sanders et al. (2016) presented the most comprehensive theory, generalizing Erkal et al.'s work to include impacts on streams of varying densities on inclined and eccentric orbits using an action-angle approach.

    The significance of an impact is measured by the change in momentum imparted to the stream stars by the perturber. To first order, we follow the developments from collision theory as laid out in Galactic Dynamics (Binney \& Tremaine) and similar works by John Taylor. In the simplest case, the perturber is considered unperturbed during the interaction, and thus has constant velocity. We can consider the most basic scenario of one point mass impacting another. A useful approximation for the change in momentum is the force at the closest approach multiplied by an estimate of the interaction time: \begin{equation} \Delta v \approx \text{Force} \times \text{interaction time} = \frac{GM}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, \end{equation} where $M$ is the mass of the perturber, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. We compute this for the five closest approaches of all globular clusters throughout each of the 50 simulations. Details on how this computation was performed are given in Appendix~\ref{sec:reconstruction}. Ultimately, we obtained a sample of 250 impacts and compared the impact geometry of those that caused gaps with those that did not.

@@

    One striking result is that more than half of the observed gaps are caused by NGC 2808, which had a flyby about 200 Myr ago. In fact, 44 of the 50 simulations show a gap from this interaction. Upon examining their morphology (Appendix~\ref{sec:gallery_of_gaps}), we observe that these gaps are consistently similar in size and location within the leading tail across most simulations. This is no coincidence. As demonstrated by Sanders and Erkal, gap creation is a dispersion phenomenon, where stars affected by a perturber acquire slightly different orbital frequencies, causing them to gradually spread apart over time. Thus, smaller gaps are indicative of more recent encounters.

    It is also expected that a recent flyby would appear in most simulations, as uncertainties in a cluster's precise location increase as we look further back in time. In general, after enough time passes, different orbital solutions \textit{phase mix}, meaning a body could be found anywhere in its phase space given its Hamiltonian and initial conditions. The characteristic phase-mixing time scale is proportional to its orbital period. However, since the encounter between NGC~2808 and Palomar~5's stream occurred less than one orbital period ago for either cluster, the orbital solutions remain relatively similar, causing the flyby to appear consistently in most simulations.

    Next, we observe that most gaps occur in the leading tail. We believe this to be a coincidence, as we expect there to be no physical motivator to cause more gaps to occur in one tail than another. First, stars escape from the globular cluster's L1 and L2 Lagrange points at roughly equal rates, leading the same size and number of stars per tail. The two main differences each tail is that the particles find themselves at different locations on the orbital phase at a given point and time. The second is that the leading tail is about 400~pc closer to the Galactic center than the trailing tail. However, this difference is negligible compared to the orbital radius of 8-12kpc. As such, we expect the likelihood of a gap appearing in either tail to be equal, with the presence of one gap not influencing the probability of another.

    Yet, the observation of only 8 gaps in the trailing tail out of 73 total gaps is inconsistent with this expectation. However, 44 of these gaps are associated with NGC~2808. As shown in Fig.\ref{fig:histogram_impact_time}, all of these gaps result from the same recent encounter. We thus discard NGC~2808 for the current analysis, as its gaps violate the assumption of independent events. After excluding these, we are left with 21 gaps in the leading tail and 8 in the trailing tail across the remaining simulations. By integrating the binomial distribution, we compute that the probability of having 8 successes or fewer given 29 trials is approximately 1.2\%. While this outcome is unlikely it is not impossible. Additionally, by studying Fig.~\ref{fig:histogram_impact_time}, we notice that, although in much smaller numbers compared to NGC~2808, other pertubers cause gaps at the same time accross different simulations. These could also be violating the assumption of independent events.

@@

      With the perturbers identified, we perform statistics in the pursuit of understanding what conditions are necessary for a globular cluster and a stream to have in order for a gap to form. We turn to impact theory, which in its simplest form as presented in works such as Binney and Tremaine, John Taylor, and Galaxiesbook, considers two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is often employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption greatly simplifies the computation.

>>>>

\newcommand{\paola}[1]{\textcolor{magenta}{{#1}}}
\newcommand{\salvatore}[1]{\textcolor{green}{{#1}}}
\newcommand{\marcook}[1]{\textcolor{green}{{#1}}}
\newcommand{\marcoko}[1]{\textcolor{red}{{#1}}}
\newcommand{\qmar}[1]{\textcolor{brown}{{#1}}}

@@

  As laid out in \citet{2021PrPNP.11903865A}'s review, despite significant progress, the true nature of dark matter remains elusive and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and galactic scales. A key focus today is detecting dark matter on sub-galactic scales, because at these scales different cosmological scenarios predict different lumpiness of the halos. For instance, \citet{2008ApJ...679.1260M} compared the Via Lactea simulations to some that they made themselves and showed that you can have a large variation for the number of dark matter subhalos within a galaxy. \citet{2019Galax...7...81Z} provides a review for the different proposed mechanisms of forming dark matter haloes from perturbation theory and deviations from a uniformly dense medium and also disccses techniques and limitations for modelign such physical phenomena. \citet{2019Galax...7...81Z} rightly states that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive. The possibilities for sub-galactic behaviour and structure of dark matter are vaste. However, as proposed by Carlberg in 2012, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. They are one of the evidences for the hierarchical nature of the galaxy formation as predicted by the $\Lambda CDM$ model \citep{white91}. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2006ApJ...637L..29B, 2009ApJ...693.1118G}. The review of Bonaca properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{2016A&A...595A...1G}.

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{2023MNRAS.520.5225M} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose memebers have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020(???) reports that the Phenoix stream is dynamically consistent from a globular cluster yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times.

  \citet{2022MNRAS.516.5331M} showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, \citet{2022A&A...667A.112V} demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are mant cluster in the internal regional of the galaxy, a region overwhich \texttt{streamfinder} was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, \citet{2018MNRAS.474.2479B} demonstrates that internal dynamics preferentially kicks out small stars first, which have lower luminosities. Earlier escape times places larger distances to the cluster and make detection more difficult.

  One of the interests in studying stellar streams, is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. ompared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams that yield the most information about the galactic potential. However, the use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, johnston05, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. Nonetheless, a recent goal of stellar stream astronomy was realized by \citet{2024ApJ...967...89I} constructed the most accurate model of the Milky Way to date based on a set of about 90 streams.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. Carlberg 2013 further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{2019ApJ...880...38B} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}.

  Our study focuses on the influence of globular clusters on stellar streams, specifically examining whether globular clusters could produce gaps similar to those expected from dark matter subhalos. If globular clusters do indeed create such gaps, this factor must be considered when using stellar streams to detect dark matter. An additional and unique aspect of our study is that we focus on modeling real stellar streams. We focus on the Palomar 5 stream, which is well-studied, has observable gaps \citep{2020ApJ...889...70B}, and is highly visible due to its position above the galactic disk. We present a result from one of our simulations in Fig.~\ref{fig:stream_on_sky}, and the following sections describe our modeling of the stream, the gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations affecting the stream.

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by Baumgardt and Vasiliev et al. We handle the uncertainties on the measured phase space parameters through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions and includes a covariance term between the proper motions---uncertainties for the right ascension, declination, and the characterisitcs radii are not available. We repeate simulation the creation of Palomar 5's tidail tails 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. For the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

    We convert the initial conditions from sky coordinates into a galactocentric reference frame. We use the local standard of rest and the peculiary motion of the sun as reported by Sch\"onrich 2010. The circular velocity at the solar radius was set to: $v_{\text{LSR}} = 240$~km~s$^{-1}$. The peculiar velocity of the sun was set to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$. The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from Chen et al 2001 and the galacto-radial distance was taken from Reid et al 2014. These transformations were performed using the \texttt{astropy} \citep{2013A&A...558A..33A}.

    For the galactic potential, we used the second static model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table 1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic model.

    The primary methodological departure from \citet{2023A&A...673A..44F} is that now we include globular cluster interactions. Instead of treating the clusters as point masses that only experience the gravitational potential of the galaxy, we now compute N-body interactions between the clusters during the backward integration over 5 billion years. The equations of motion for the globular cluster system are thus: 
    \begin{equation}
      \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
    \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galacto-centric position, the index $i$ indicates a globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates. The index $j$ indicates the other globular cluster that are summed over where: $m_j$ is its mass, $b_j$ is its plummer scale radius, and $r_j$ is it's galacto-centric position. $\Phi$ is the axis-symmteric potential of the model used for the Milky Way.

    At their computed positions 5 Gyr ago, each globular cluster was assigned 100,000 particles sampled from a Plummer distribution to represent its stellar content. For each set of initial conditions, the particle positions were resampled to account for uncertainties in mass, which is an improvement from \citet{2023A&A...673A..44F}. Since globular clusters are older than 8 Gyr, their most massive stars have already gone supernova, leaving primarily low-mass stars, typically under one solar mass (CITATION?). Given that clusters usually have masses between $10^5-10^6$ solar masses, 100~000 particles is sufficient to represent the distribution of tidal debris.

@@

    The integration was performed using a leapfrog integrator (add citation). The timestep size was chosen to be based on two critera. First, to ensure the conservation of energy to XXX. The second, is the drift in position due to numerical error over 5 Gyr was negligible compared to the scale radius of the globular clusters, which is about 5~pc. Further details on the timestep selection are discussed in the appendix.

@@

    At the end of the simulations, some gaps in the stellar stream are so prominent that they can be identified visually without any sophisticated quantitative analysis. The analysis of the morphology of the tidal tails of Palomar~5 is carried out by comparing simulations with globular clusters (called \texttt{full}) and without them (called \texttt{reference}). A preliminary evaluation is performed visually, followed by a quantitative analysis of the differences between the density profiles in the two mentioned cases.

    % We initially sought an automated, quantitative approach from signal and image processing that could reliably detect all gaps and evaluate their significance. Unfortunately, despite extensive efforts, an ideal method remained elusive. As a result, we opted for a simpler approach: a combination of visual inspection aided by comparative analysis.

    The \texttt{reference} simulations used the same 50 sets of initial conditions but excluded interactions with globular clusters, except for the host cluster, Palomar 5. In other words, the gravitational influence of all other clusters on the star particles within the stream was omitted. By comparing the morphological differences between these \texttt{reference} simulations and the \texttt{full} simulations, we could more effectively identify and assess the impact of globular clusters on the stream's structure. Details on gap identification are provided in Appendix~\ref{sec:gap_detection}, and the identification of corresponding perturbing globular clusters is detailed in Appendix~\ref{sec:Perturber_Identification}.

    Comparisons between the \texttt{reference} and \texttt{full} simulations were performed in the tail coordinate system. This system allows for consistent morphological comparisons by aligning the stream with the cluster's orbit. The coordinate system is based on the work of \citet{2004AJ....127.2753D} and is detailed in Appendix~\ref{appendix:TailCoordinates}. Briefly, the x' coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead or behind the cluster. The y' coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer.

    Once the gaps and perturbers were identified, we sought to characterize the properties of the globular clusters that caused gaps and differentiate them from those that did not. \citet{2013ApJ...775...90C} initiated the theory of gap creation in stellar streams due to gravitational flybys. This analysis was extended both analytically and numerically by \citet{2015MNRAS.450.1136E}, who studied uniform-density streams on circular orbits. \citet{2016MNRAS.457.3817S} presented the most comprehensive theory, generalizing Erkal et al.'s work to include impacts on streams of varying densities on inclined and eccentric orbits using an action-angle approach.

    The significance of an impact is measured by the change in momentum imparted to the stream stars by the perturber. A useful approximation for the change in momentum is the force at the closest approach multiplied by an estimate of the interaction time: \begin{equation} \Delta v \approx \text{Force} \times \text{interaction time} = \frac{GM}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, \end{equation} where $M$ is the mass of the perturber, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. We compute this for the five closest approaches of all globular clusters throughout each of the 50 simulations. Details on how this computation was performed are given in Appendix~\ref{sec:reconstruction}. Ultimately, we obtained a sample of 250 impacts and compared the impact geometry of those that caused gaps with those that did not.

@@

    NGC~2808 impacts Palomar 5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position in each case. The gap size aligns with gap formation as a dispersion process, where stars gradually spread apart due to slight differences in orbital frequencies. Smaller gaps indicate more recent encounters. Since this interaction occurred less than one orbital period ago, the orbital solutions remain similar, producing consistent results across simulations.

    Most gaps appear in the leading tail, which we consider coincidental, as stars escape the cluster's L1 and L2 points at similar rates, and the small distance difference between tails' galactocentric distance is negligible. We expect gaps to be equally likely in both tails. However, only 8 of the 73 gaps occur in the trailing tail, with 44 linked to NGC2808's flyby. After removing these, there are 21 gaps in the leading tail and 8 in the trailing. By the binomial distribution, the probability of observing up to 8 successes in 29 trials, given a 50\% chance of success, is 1.2\%—unlikely, but possible. Additionally, other perturbers impact at consistent times, which may also violate the assumption of independent events, as seen with NGC2808.

@@

      With the perturbers identified, we perform statistics in the pursuit of understanding what conditions are necessary for a globular cluster and a stream to have in order for a gap to form. We turn to impact theory, which in its simplest form as presented in works such as \citet{2008gady.book.....B}, John Taylor, and Galaxiesbook, considers two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is often employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption greatly simplifies the computation.
